Closure-Free Functional Programming in a Two-Level Type Theory

### 2024-02-29 newpaper/paper.tex
stuff

Modern functional programming supports many convenient abstractions. These often
come with significant runtime overheads. Sometimes the overheads are acceptable,
but in other cases compiler optimization is crucial. Monads in Haskell is an
example for the latter. Even the $\Reader$ monad, which is one of the simplest
in terms of implementation, yields large overheads when compiled without
optimizations. Consider the following:
\begin{alignat*}{3}
  &\vf :: \Int \to \Reader\,\Bool\,\Int \\
  &\vf\,\vx = \mdo\{\vb \fro \msf{ask}; \bif \vx\, \bthen \return (\vx + 10)\, \belse \return (\vx + 20)\}
\end{alignat*}
With optimizations enabled, GHC compiles this roughly to the code below:
\begin{alignat*}{3}
  &\vf :: \Int \to \Bool \to \Int \\
  &\vf = \lam \vx\,\vb.\, \bif \vb\, \bthen \vx + 10\, \belse \vx + 20
\end{alignat*}
Without optimizations we roughly get:
\begin{alignat*}{3}
  &\vf = \lam \vx.\,(\bind)\,\msf{MonadReaderDict}\,\msf{ask}\,(\lam \vb.\,\bif \vb\\
  &\ind \bthen\return\,\msf{MonadReaderDict}\,(\vx + 10)\\
  &\ind \belse\hspace{0.25em}\return\,\msf{MonadReaderDict}\,(\vx + 20))
\end{alignat*}
Here, $\msf{MonadReaderDict}$ is a runtime dictionary, containing the methods of
the $\Monad$ instance for $\Reader$, and $(\bind)\,\msf{MonadReaderDict}$ is a
field projection. Here, a runtime closure will be created for the $\lam \vb.\,...$
function, and $(\bind)$, $\msf{ask}$ and $\return$ will create additional dynamic
closures.

@@

GHC's optimization efforts are respectable, and it has gotten quite adept over
its long history of development. However, there is no strong guarantee that
certain optimizations will happen. Control over optimizations remains tricky,
fragile and non-compositional. \texttt{INLINE} and \texttt{REWRITE} pragmas can
be used to control code generation \cite{ghcdocs}, but without strong
guarantees, and their advanced usage requires knowledge of GHC internals. For
example, correctly specifying the \emph{ordering} of certain rule applications
is often needed. We also have to care about formal function arities. Infamously,
the function composition operator is defined as $(.)\,\vf\,\vg = \lam \vx \to
\vf\,(\vg\,\vx)$ in the base libraries, instead of as $(.)\,\vf\,\vg\,\vx =
\vf\,(\vg\,\vx)$, to get better inlining behavior \cite{ghcbase}. Tellingly, it
is common practice in high-performance Haskell programming to visually review
GHC's optimized code output.

@@

A CFTT program is a mixture of object-level and meta-level top-level definitions
and declarations. \textbf{Staging} means running all metaprograms in splices and
inserting their output into object code, keeping all object-level top entries
and discarding all meta-level ones. Thus, staging takes a CFTT program as input,
and produces output which is purely in the object-level fragment, with no
metatypes and metaprograms remaining. The output is guaranteed to be well-typed.

Let us look at some basic staging examples. Recall the meta-level identity
function; it can be used at the object-level too, by applying it to quoted
terms:
  \[ \letdef \vn : \Nat := \spl(\id\,\ql 10 + 10 \qr); ... \]
Here, $\id$ is used at type $\Lift \Nat$. During staging, the expression in the
splice is evaluated, so we get $\spl\ql 10 + 10 \qr$, which is definitionally
the same as $10 + 10$, which is our staging output here. Boolean
short-circuiting is another basic use-case:
\begin{alignat*}{3}
  &\mit{and} : \Up\Bool \to \Up\Bool \to \Up\Bool\\
  &\mit{and}\,\vx\,\vy = \ql\case \spl \vx\,\of \True \to \spl \vy; \False \to \False\qr
\end{alignat*}
Since the $y$ expression is inlined under a $\case$ branch at every use site, it
is only computed at runtime when $x$ evaluates to $\True$. In many situations,
staging can be used instead of laziness to implement short-circuiting, and often
with better runtime performance, avoiding the overhead of thunking. Consider the
$\map$ function now:
\begin{alignat*}{3}
  & \hspace{-5em}\mathrlap{\map : \{\vA\,\vB : \VTy\} \to (\Up \vA \to \Up \vB) \to \Up (\List\,\vA) \to \Up(\List\,\vB)}\\
  & \hspace{-5em}\mathrlap{\map\,\vf\,\vas = \ql\letrec \go\,\vas := \case \vas\,\of}\\
  & \hspace{-5em}\ind\ind\ind\ind\ind \ind\Nil             &&\to \Nil;\\
  & \hspace{-5em}\ind\ind\ind\ind\ind \ind\Cons\,\va\,\vas &&\to \Cons\,\spl(\vf\,\ql \va \qr)\,(\go\,\vas);\\
  & \hspace{-5em}\ind\ind\ind\ind\ind\go\,\spl\vas \qr
\end{alignat*}
For example, this can be used as $\letdef f\,\vas : \List\,\Nat \to \List\,\Nat
:= \spl(\map\,(\lam \vx.\,\ql \spl \vx + 10 \qr)\,\qt{\vas}) $.  This is staged to a
recursive definition where the mapping function is inlined into the $\Cons$ case
as $\Cons\,\va\,\vas \to \Cons\,(\va + 10)\,(\go\,\vas)$.  Note that $\map$ has to
abstract over value types, since lists can only contain values, not
functions. Also, the mapping function has type $\Up \vA \to \Up \vB$, instead of
$\Up (\vA \to \vB)$. The former type is often preferable to the latter in staging;
the former is a metafunction with useful computational content, while the latter
is merely a black box that computes object code. If we have $\vf : \Up(\vA \to \vB)$,
and $\vf$ is staged to $\ql \lam \vx.\,\vt\qr$, then $\spl \vf\,\vu$ is staged to an
undesirable ``administrative'' $\beta$-redex $(\lam \vx.\,\vt)\,\vu$.

@@

The solution is to use a monad which extends $\MTy$ with the ability to
freely generate object-level code and introduce object binders.
\begin{alignat*}{3}
  & \newtype \Gen\,(\vA : \MTy) = \Gen\,\{\unGen : \{\vR : \Ty\} \to (\vA \to \Up \vR) \to \Up \vR\}
\end{alignat*}
This is a monad in $\MTy$ in a standard sense:
\begin{alignat*}{3}
  &\mathrlap{\instance \Monad\,\Gen\,\where}\\
  &\ind \return \va        &&= \Gen \dlr \lam \vk.\,\vk\,\va\\
  &\ind \mit{ga} \bind \vf &&= \Gen \dlr \lam \vk.\,\unGen\,\mit{ga}\,(\lam \va.\,\unGen\,(\vf\,\va)\,\vk))
\end{alignat*}
From now on, we reuse Haskell-style type classes and $\mbf{do}$-notation in
CFTT. We will use type classes in an informal way, without precisely specifying
how they work. However, type classes are used in essentially the same way in the
Agda and Haskell implementations of the paper, with modest technical
differences.

From the $\Monad$ instance, the $\Functor$ and $\Applicative$ instances can be
also derived. We reuse $({\fmap})$ and $({\ap})$ for applicative notation as well.
An inhabitant of $\Gen\,\vA$ can be viewed as an action whose effect is to produce
some object code. $\Gen$ can be only ``run'' when the result type is an object
type:
\begin{alignat*}{3}
  &\runGen : \Gen\,(\Up \vA) \to \Up \vA\\
  &\runGen\,\vma = \unGen\,\vma\,\id
\end{alignat*}
We can let-bind object expressions in $\Gen$:
\begin{alignat*}{3}
  & \gen : \Up \vA \to \Gen\,(\Up \vA) \\
  & \gen\,\va = \Gen \dlr \lam \vk.\,\ql \letdef \vx : \vA := \spl \va; \spl(\vk\,\ql \vx \qr) \qr
\end{alignat*}
And also recursive definitions of computations:
\begin{alignat*}{3}
  & \genRec : \{\vA : \CTy\} \to (\Up \vA \to \Up \vA) \to \Gen\,(\Up \vA) \\
  & \genRec\,\vf = \Gen \dlr \lam \vk.\,\qt{\letrec \vx : \vA := \spl(\vf\,\qt{\vx}); \spl(\vk\,\qt{\vx})}
\end{alignat*}
Now, using do-notation, we may write $\mdo \{\vx \fro \gen\,\qt{10 + 10}; \vy \fro
\gen\,\qt{20 + 20}\};\return \qt{\vx + \vy}\}$, for a $\Gen\,(\Up \Nat)$
action. Running this with $\runGen$ yields $\qt{\letdef \vx := 10 + 10; \letdef \vy
  := 20 + 20; \vx + \vy}$. We can also define a ``safer'' binding-time improvement for
products, using let-insertion:
\begin{alignat*}{6}
  &\hspace{-2em}\mathrlap{\mup : \Up (\vA,\,\vB) \to \Gen\,(\Up \vA,\,\Up \vB)            \hspace{5em} \mdown : \Gen\,(\Up \vA,\, \Up \vB) \to \Up(\vA,\,\vB)} \\
  &\hspace{-2em}\mup\,\vx = \mdo &&\vx \fro \gen\,\vx                                    && \ind\ind \mdown\,\vx = \mdo &&(\va,\,\vb) \fro \vx\\
  &\hspace{-2em}               &&\return (\qt{\fst\,\spl \vx},\,\qt{\snd\,\spl \vx})     &&                           &&\return \qt{(\spl \va,\, \spl \vb)}
\end{alignat*}

@@

We demonstrate this for lists first. An object-level $\mbf{case}$ on lists introduces
two points where code generation can continue. We define a metatype which gives
us a ``view'' on these points:
\[ \data \msf{SplitList}\,\vA = \Nil'\,|\,\Cons'\,(\Up \vA)\,(\Up (\List\,\vA)) \]
We can generate code for a case split, returning a view on it:
\begin{alignat*}{3}
  &\mit{split} : \Up (\List\,\vA) \to \Gen\,(\msf{SplitList}\,\vA)\\
  &\mit{split}\,\vas = \Gen \dlr \lam \vk.\,\qt{\case \spl \vas\,\of \Nil \to \spl(\vk\,\Nil'); \Cons\,\va\,\vas \to \spl(\vk\,(\Cons'\,\qt{a}\,\qt{\vas}))}
\end{alignat*}
Now, in any $\MonadGen$, assuming $\vas : \Up(\List\,\vA)$, we may write
\begin{alignat*}{3}
  &\mdo \{\mit{sp} \fro \liftGen\;(\!\mit{split}\,\vas);(\case \mit{sp}\,\of \Nil' \to ...;\Cons'\,\va\,\vas \to ...)\}
\end{alignat*}
This can be generalized to splitting on any object value. In the Agda and
Haskell implementations, we overload $\mit{split}$ with a class similar to the
following:
\begin{alignat*}{3}
  & \hspace{-2em}\mathrlap{\tyclass\,\mit{Split}\,(\vA : \VTy)\,\where}\\
  & \hspace{-2em}\ind \mit{SplitTo} &&: \MTy \\
  & \hspace{-2em}\ind \mit{split}   &&: \Lift \vA \to \Gen\,\mit{SplitTo}
\end{alignat*}
In a native implementation of CFTT it may make sense to extend do-notation, so
that we elaborate $\mbf{case}$ on object values to an application of the
appropriate $\mit{split}$ function. We adopt this in the rest of the paper, so
when working in a $\MonadGen$, we can write $\case \vas\,\of \Nil \to ...;
\Cons\,\va\,\vas \to ...$, binding $\va : \Up \vA$ and $\vas : \Up(\List\,\vA)$
in the $\Cons$ case.

@@

At this point, it makes sense to aim for a monad transformer library where
binding-time improvement is defined compositionally, by recursion on the
transformer stack. The base case is the following:
\begin{alignat*}{3}
  & \hspace{-2em}\mathrlap{\newtype \Identity\,\vA := \Identity\,\{\runIdentity : \vA\} }\\
  & \hspace{-2em}\mathrlap{\instance \Improve\,\Identity\,\Gen\,\where}\\
  & \hspace{-2em}\ind \mup\,\vx   &&= \return \qt{\runIdentity\,\spl \vx}\\
  & \hspace{-2em}\ind \mdown\,\vx &&= \unGen\,\vx \dlr \lam \va.\,\qt{\Identity\,\spl \va}
\end{alignat*}
We also recover the object-level $\Maybe$ as $\MaybeT\,\Identity$, from the following $\MaybeT$:
\[ \newtype \MaybeT\,(\vM : \VTy \to \Ty)\,(\vA : \VTy) := \MaybeT\,\{\runMaybeT : \vM\,(\Maybe\,\vA)\} \]
With this, improvement can be generally defined for $\MaybeT$:
\begin{alignat*}{3}
  &\instance \Improve\,\vF\,\vM \RA \Improve\,(\MaybeT\,\vF)\,(\MaybeT_\M\,\vM)\,\where\\
  &\ind \mup\,\vx = \MaybeT_\M \dlr \mdo\\
  &\ind\ind \vma \fro \mup\,\qt{\runMaybeT\,\spl \vx}\\
  &\ind\ind \case \vma\,\of \Nothing \to \return \Nothing_\M\\
  &\hspace{6.5em}         \Just\,\va \hspace{1.15em}\to \return (\Just_\M\,\va)\\
  &\ind \mdown\,(\MaybeT_\M\,\vx) = \qt{\MaybeT\,\spl(\mdown \dlr \vx \bind \lam\case\\
  &\ind\ind\Nothing_\M \to \return \qt{\Nothing}\\
  &\ind\ind\Just_\M\,\va\hspace{1.15em} \to \return \qt{\Just\,\spl \va})}
\end{alignat*}
In the $\mbf{case}$ in $\mup$, we use our syntax sugar for matching on a
$\Maybe$ value inside an $\vM$ action. This is legal, since we know from the
$\Improve\,\vF\,\vM$ assumption that $\vM$ is a $\MonadGen$. In $\mdown$ we also use
$\lam\case ...$ to shorten $\lam \vx.\,\case \vx\,\of ...$.

@@

Let us look at the staging output. We omit newtype wrappers and use nested
pattern matching on pairs, but otherwise this is exactly the code that we get in
our Agda implementation.
\begin{alignat*}{4}
  &  \letrec &&\vf : \Tree\,\Nat \to \StateT\,(\List\,\Nat)\,(\MaybeT\,\Identity)\,(\Tree\,\Nat) \\
  &          &&\vf := \lam \vt\,\vns.\, \case\,\vt\,\of\\
  &          &&\ind \Leaf \to \Just\,(\Leaf,\,\vns) \\
  &          &&\ind \Node\,\vn\,\vl\,\vr \to \case (\vn == 0)\,\of\\
  &          &&\ind \ind \True \hspace{0.25em}\to \Nothing \\
  &          &&\ind \ind \False \to \\
  &          &&\ind \ind \ind \letdef \msf{joinNothing} := \lam \_.\,\Nothing;\\
  &          &&\ind \ind \ind \letdef \msf{joinJust} := \lam \vn\,\vns.\,\case\,\vf\,\vl\,\vns\,\of\\
  &          &&\ind \ind \ind \ind \ind \ind \Nothing\hspace{0.9em} \to \Nothing\\
  &          &&\ind \ind \ind \ind \ind \ind \Just\,(\vr,\,\vns) \to \case\,\vf\,\vr\,\vns\,\of\\
  &          &&\ind \ind \ind \ind \ind \ind \ind \Nothing\hspace{.95em} \to \Nothing\\
  &          &&\ind \ind \ind \ind \ind \ind \ind \Just\,(\vr,\,\vns) \to \Just\,(\Node\,\vn\,\vl\,\vr,\,\vns);\\
  &          &&\ind \ind \ind \case \vns\,\of\\
  &          &&\ind \ind \ind \ind \Nil \hspace{2.65em}\to \msf{joinJust}\,\vn\,\vns\\
  &          &&\ind \ind \ind \ind \Cons\,\vn\,\vns \to \msf{joinJust}\,\vn\,\vns
\end{alignat*}
The only flaw in this code is the $\msf{joinNothing}$, which is never called,
becase we joined an action that never throws an error. But this is also
very easy to clean up after staging.
\subsection{Discussion}

@@

Let us write $\Uprod = \List\,\VTy$ and $\Elprod : \Uprod \to \MTy$ for a
universe of finite products of value types. Concretely, in CFTT and the Agda
implementation, we assume the following:
\\\\
\noindent \textbf{Axiom (generativity).}\textit{ Every $\vf : \Elprod\,\vA \to \USOP$ is constant. }
\\\\
We remark that there is no risk of staging getting stuck on this axiom, because
propositional equality proofs get erased, as we described in Section
\ref{sec:semantics-of-staging}.

>>>>

Modern functional programming supports many convenient abstractions. These often
come with significant runtime overheads. Sometimes the overheads are acceptable,
but in other cases compiler optimization is crucial. Monads in Haskell is an
example for the latter. Even the $\Reader$ monad, which is one of the simplest
in terms of implementation, yields large overheads when compiled without
optimizations. Consider the following:
\begin{alignat*}{3}
  &\vf :: \Int \to \Reader\,\Bool\,\Int \\
  &\vf\,\vx = \mdo\{\vb \fro \msf{ask}; \bif \vx\, \bthen \return (\vx + 10)\, \belse \return (\vx + 20)\}
\end{alignat*}
With optimizations enabled, GHC compiles this roughly to the code below:
\begin{alignat*}{3}
  &\vf :: \Int \to \Bool \to \Int \\
  &\vf = \lam \vx\,\vb.\, \bif \vb\, \bthen \vx + 10\, \belse \vx + 20
\end{alignat*}
Without optimizations we roughly get:
\begin{alignat*}{3}
  &\vf = \lam \vx.\,(\bind)\,\msf{MonadReaderDict}\,\msf{ask}\,(\lam \vb.\,\bif \vb\\
  &\ind \bthen\return\,\msf{MonadReaderDict}\,(\vx + 10)\\
  &\ind \belse\hspace{0.25em}\return\,\msf{MonadReaderDict}\,(\vx + 20))
\end{alignat*}
Here, $\msf{MonadReaderDict}$ is a runtime dictionary, containing the methods of
the $\Monad$ instance for $\Reader$, and $(\bind)\,\msf{MonadReaderDict}$ is a
field projection. Here, a runtime closure will be created for the $\lam \vb.\,...$
function, and $(\bind)$, $\msf{ask}$ and $\mbf{return}$ will create additional dynamic
closures.

@@

GHC's optimization efforts are respectable, and it has gotten quite adept over
its long history of development. However, there is no strong guarantee that
certain optimizations will happen. Control over optimizations remains tricky,
fragile and non-compositional. \texttt{INLINE} and \texttt{REWRITE} pragmas can
be used to control code generation \cite{ghcdocs}, but without strong
guarantees, and their advanced usage requires knowledge of GHC internals. For
example, correctly specifying the \emph{ordering} of certain rule applications
is often needed. We also have to care about formal function arities. Infamously,
the function composition operator is defined as $(.)\,\vf\,\vg = \lam \vx \to
\vf\,(\vg\,\vx)$ in the base libraries, instead of as $(.)\,\vf\,\vg\,\vx =
\vf\,(\vg\,\vx)$, to get better inlining behavior \cite{ghcbase}. It is common
practice in high-performance Haskell programming to visually review GHC's
optimized code output.

@@

A CFTT program is a mixture of object-level and meta-level top-level definitions
and declarations. \emph{Staging} means running all metaprograms in splices and
inserting their output into object code, keeping all object-level top entries
and discarding all meta-level ones. Thus, staging takes a CFTT program as input,
and produces output which is purely in the object-level fragment, with no
metatypes and metaprograms remaining. The output is guaranteed to be well-typed.

Let us look at some basic staging examples. Recall the meta-level identity
function; it can be used at the object-level too, by applying it to quoted
terms:
  \[ \letdef \vn : \Nat := \spl(\id\,\ql 10 + 10 \qr); ... \]
Here, $\id$ is used at type $\Lift \Nat$. During staging, the expression in the
splice is evaluated, so we get $\spl\ql 10 + 10 \qr$, which is definitionally
the same as $10 + 10$, which is our staging output here. Boolean
short-circuiting is another basic use-case:
\begin{alignat*}{3}
  &\mit{and} : \Up\Bool \to \Up\Bool \to \Up\Bool\\
  &\mit{and}\,\vx\,\vy = \ql\case \spl \vx\,\of \True \to \spl \vy; \False \to \False\qr
\end{alignat*}
Since the $y$ expression is inlined under a $\case$ branch at every use site, it
is only computed at runtime when $x$ evaluates to $\True$. In many situations,
staging can be used instead of laziness to implement short-circuiting, and often
with better runtime performance, avoiding the overhead of thunking. Consider the
$\map$ function now:
\begin{alignat*}{3}
  & \hspace{-5em}\mathrlap{\map : \{\vA\,\vB : \VTy\} \to (\Up \vA \to \Up \vB) \to \Up (\List\,\vA) \to \Up(\List\,\vB)}\\
  & \hspace{-5em}\mathrlap{\map\,\vf\,\vas = \ql\letrec \go\,\vas := \case \vas\,\of}\\
  & \hspace{-5em}\ind\ind\ind\ind\ind \ind\Nil             &&\to \Nil;\\
  & \hspace{-5em}\ind\ind\ind\ind\ind \ind\Cons\,\va\,\vas &&\to \Cons\,\spl(\vf\,\ql \va \qr)\,(\go\,\vas);\\
  & \hspace{-5em}\ind\ind\ind\ind\ind\go\,\spl\vas \qr
\end{alignat*}
For example, this can be used as $\letdef \vf\,\vas : \List\,\Nat \to \List\,\Nat
:= \spl(\map\,(\lam \vx.\,\ql \spl \vx + 10 \qr)\,\qt{\vas}) $.  This is staged to a
recursive definition where the mapping function is inlined into the $\Cons$ case
as $\Cons\,\va\,\vas \to \Cons\,(\va + 10)\,(\go\,\vas)$.  Note that $\map$ has to
abstract over value types, since lists can only contain values, not
functions. Also, the mapping function has type $\Up \vA \to \Up \vB$, instead of
$\Up (\vA \to \vB)$. The former type is often preferable to the latter in staging;
the former is a metafunction with useful computational content, while the latter
is merely a black box that computes object code. If we have $\vf : \Up(\vA \to \vB)$,
and $\vf$ is staged to $\ql \lam \vx.\,\vt\qr$, then $\spl \vf\,\vu$ is staged to an
undesirable ``administrative'' $\beta$-redex $(\lam \vx.\,\vt)\,\vu$.

@@

The solution is to use a monad which extends $\MTy$ with the ability to
freely generate object-level code and introduce object binders.
\begin{alignat*}{3}
  & \newtype \Gen\,(\vA : \MTy) = \Gen\,\{\unGen : \{\vR : \Ty\} \to (\vA \to \Up \vR) \to \Up \vR\}
\end{alignat*}
This is a monad in $\MTy$ in a standard sense:
\begin{alignat*}{3}
  &\mathrlap{\instance \Monad\,\Gen\,\where}\\
  &\ind \return \va        &&= \Gen \dlr \lam \vk.\,\vk\,\va\\
  &\ind \mit{ga} \bind \vf &&= \Gen \dlr \lam \vk.\,\unGen\,\mit{ga}\,(\lam \va.\,\unGen\,(\vf\,\va)\,\vk))
\end{alignat*}
From now on, we reuse Haskell-style type classes and $\mbf{do}$-notation in
CFTT. We will use type classes in an informal way, without precisely specifying
how they work. However, type classes are used in essentially the same way in the
Agda and Haskell implementations, with modest technical differences.

From the $\Monad$ instance, the $\Functor$ and $\Applicative$ instances can be
also derived. We reuse $({\fmap})$ and $({\ap})$ for applicative notation as well.
An inhabitant of $\Gen\,\vA$ can be viewed as an action whose effect is to produce
some object code. $\Gen$ can be only ``run'' when the result type is an object
type:
\begin{alignat*}{3}
  &\runGen : \Gen\,(\Up \vA) \to \Up \vA\\
  &\runGen\,\vma = \unGen\,\vma\,\id
\end{alignat*}
We can let-bind object expressions in $\Gen$:
\begin{alignat*}{3}
  & \gen : \{\vA : \Ty\} \to \Up \vA \to \Gen\,(\Up \vA) \\
  & \gen\,\va = \Gen \dlr \lam \vk.\,\ql \letdef \vx : \vA := \spl \va; \spl(\vk\,\ql \vx \qr) \qr
\end{alignat*}
And also recursive definitions of computations:
\begin{alignat*}{3}
  & \genRec : \{\vA : \CTy\} \to (\Up \vA \to \Up \vA) \to \Gen\,(\Up \vA) \\
  & \genRec\,\vf = \Gen \dlr \lam \vk.\,\qt{\letrec \vx : \vA := \spl(\vf\,\qt{\vx}); \spl(\vk\,\qt{\vx})}
\end{alignat*}
Now, using do-notation, we may write $\mdo \{\vx \fro \gen\,\qt{10 + 10}; \vy \fro
\gen\,\qt{20 + 20}\};\return \qt{\vx + \vy}\}$, for a $\Gen\,(\Up \Nat)$
action. Running this with $\runGen$ yields $\qt{\letdef \vx := 10 + 10; \letdef \vy
  := 20 + 20; \vx + \vy}$. We can also define a ``safer'' binding-time improvement for
products, using let-insertion:
\begin{alignat*}{6}
  &\hspace{-2em}\mathrlap{\mup : \Up (\vA,\,\vB) \to \Gen\,(\Up \vA,\,\Up \vB)            \hspace{5em} \mdown : \Gen\,(\Up \vA,\, \Up \vB) \to \Up(\vA,\,\vB)} \\
  &\hspace{-2em}\mup\,\vx = \mdo &&\vx \fro \gen\,\vx                                    && \ind\ind \mdown\,\vx = \mdo &&(\va,\,\vb) \fro \vx\\
  &\hspace{-2em}               &&\return (\qt{\fst\,\spl \vx},\,\qt{\snd\,\spl \vx})     &&                           &&\return \qt{(\spl \va,\, \spl \vb)}
\end{alignat*}

@@

We demonstrate this for lists. An object-level $\mbf{case}$ on lists introduces
two points where code generation can continue. We define a metatype which gives
us a ``view'' on these points:
\[ \data \msf{SplitList}\,\vA = \Nil'\,|\,\Cons'\,(\Up \vA)\,(\Up (\List\,\vA)) \]
We can generate code for a case split, returning a view on it:
\begin{alignat*}{3}
  &\mit{split} : \Up (\List\,\vA) \to \Gen\,(\msf{SplitList}\,\vA)\\
  &\mit{split}\,\vas = \Gen \dlr \lam \vk.\,\qt{\case \spl \vas\,\of \Nil \to \spl(\vk\,\Nil'); \Cons\,\va\,\vas \to \spl(\vk\,(\Cons'\,\qt{a}\,\qt{\vas}))}
\end{alignat*}
Now, in any $\MonadGen$, assuming $\vas : \Up(\List\,\vA)$, we may write
\begin{alignat*}{3}
  &\mdo \{\mit{sp} \fro \liftGen\;(\!\mit{split}\,\vas);(\case \mit{sp}\,\of \Nil' \to ...;\Cons'\,\va\,\vas \to ...)\}
\end{alignat*}
This can be generalized to splitting on any object value. In the Agda and
Haskell implementations, we overload $\mit{split}$ with a class similar to the
following:
\begin{alignat*}{3}
  & \hspace{-2em}\mathrlap{\tyclass\,\mit{Split}\,(\vA : \VTy)\,\where}\\
  & \hspace{-2em}\ind \mit{SplitTo} &&: \MTy \\
  & \hspace{-2em}\ind \mit{split}   &&: \Lift \vA \to \Gen\,\mit{SplitTo}
\end{alignat*}
In a native implementation of CFTT it may make sense to extend do-notation, so
that we elaborate $\mbf{case}$ on object values to an application of the
appropriate $\mit{split}$ function. We adopt this in the rest of the paper, so
when working in a $\MonadGen$, we can write $\case \vas\,\of \Nil \to ...;
\Cons\,\va\,\vas \to ...$, binding $\va : \Up \vA$ and $\vas : \Up(\List\,\vA)$
in the $\Cons$ case.

@@

At this point, it makes sense to aim for a monad transformer library where
binding-time improvement is defined compositionally, by recursion on the
transformer stack. The base case is the following:
\begin{alignat*}{3}
  & \hspace{-2em}\mathrlap{\newtype \Identity\,\vA := \Identity\,\{\runIdentity : \vA\} }\\
  & \hspace{-2em}\mathrlap{\instance \Improve\,\Identity\,\Gen\,\where}\\
  & \hspace{-2em}\ind \mup\,\vx   &&= \return \qt{\runIdentity\,\spl \vx}\\
  & \hspace{-2em}\ind \mdown\,\vx &&= \unGen\,\vx \dlr \lam \va.\,\qt{\Identity\,\spl \va}
\end{alignat*}
We recover the object-level $\Maybe$ as $\MaybeT\,\Identity$, from the following $\MaybeT$:
\[ \newtype \MaybeT\,(\vM : \VTy \to \Ty)\,(\vA : \VTy) := \MaybeT\,\{\runMaybeT : \vM\,(\Maybe\,\vA)\} \]
With this, improvement can be generally defined for $\MaybeT$:
\begin{alignat*}{3}
  &\instance \Improve\,\vF\,\vM \RA \Improve\,(\MaybeT\,\vF)\,(\MaybeT_\M\,\vM)\,\where\\
  &\ind \mup\,\vx = \MaybeT_\M \dlr \mdo\\
  &\ind\ind \vma \fro \mup\,\qt{\runMaybeT\,\spl \vx}\\
  &\ind\ind \case \vma\,\of \Nothing \to \return \Nothing_\M\\
  &\hspace{6.5em}         \Just\,\va \hspace{1.15em}\to \return (\Just_\M\,\va)\\
  &\ind \mdown\,(\MaybeT_\M\,\vx) = \qt{\MaybeT\,\spl(\mdown \dlr \vx \bind \lam\case\\
  &\ind\ind\Nothing_\M \to \return \qt{\Nothing}\\
  &\ind\ind\Just_\M\,\va\hspace{1.15em} \to \return \qt{\Just\,\spl \va})}
\end{alignat*}
In the $\mbf{case}$ in $\mup$, we use our syntax sugar for matching on a
$\Maybe$ value inside an $\vM$ action. This is legal, since we know from the
$\Improve\,\vF\,\vM$ assumption that $\vM$ is a $\MonadGen$. In $\mdown$ we also use
$\lam\case ...$ to shorten $\lam \vx.\,\case \vx\,\of ...$.

@@

Let us look at the staging output. We omit newtype wrappers and use nested
pattern matching on pairs, but otherwise this is exactly the code that we get in
our Agda implementation.
\begin{alignat*}{4}
  &  \letrec &&\vf : \Tree\,\Nat \to \StateT\,(\List\,\Nat)\,(\MaybeT\,\Identity)\,(\Tree\,\Nat) \\
  &          &&\vf := \lam \vt\,\vns.\, \case\,\vt\,\of\\
  &          &&\ind \Leaf \to \Just\,(\Leaf,\,\vns) \\
  &          &&\ind \Node\,\vn\,\vl\,\vr \to \case (\vn == 0)\,\of\\
  &          &&\ind \ind \True \hspace{0.25em}\to \Nothing \\
  &          &&\ind \ind \False \to \\
  &          &&\ind \ind \ind \letdef \msf{joinNothing} := \lam \_.\,\Nothing;\\
  &          &&\ind \ind \ind \letdef \msf{joinJust} := \lam \vn\,\vns.\,\case\,\vf\,\vl\,\vns\,\of\\
  &          &&\ind \ind \ind \ind \ind \ind \Nothing\hspace{0.9em} \to \Nothing\\
  &          &&\ind \ind \ind \ind \ind \ind \Just\,(\vl,\,\vns) \to \case\,\vf\,\vr\,\vns\,\of\\
  &          &&\ind \ind \ind \ind \ind \ind \ind \Nothing\hspace{.95em} \to \Nothing\\
  &          &&\ind \ind \ind \ind \ind \ind \ind \Just\,(\vr,\,\vns) \to \Just\,(\Node\,\vn\,\vl\,\vr,\,\vns);\\
  &          &&\ind \ind \ind \case \vns\,\of\\
  &          &&\ind \ind \ind \ind \Nil \hspace{2.65em}\to \msf{joinJust}\,\vn\,\vns\\
  &          &&\ind \ind \ind \ind \Cons\,\vn\,\vns \to \msf{joinJust}\,\vn\,\vns
\end{alignat*}
The only flaw in this code is the $\msf{joinNothing}$, which is never called,
becase we joined an action that never throws an error. But this is also
very easy to clean up after staging.
\subsection{Discussion}

@@

Let us write $\Uprod = \List\,\VTy$ and $\Elprod : \Uprod \to \MTy$ for a
universe of finite products of value types. Concretely, in CFTT and the Agda
implementation, we assume the following:
\\\\
\noindent \textbf{Axiom (generativity).}\textit{ For every $\vf : \Elprod\,\vA \to \USOP$ and $\va,\,\va' : \Elprod\,\vA$,
we have that $\vf\,\va = \vf\,\va'$.}
\\\\
We remark that there is no risk of staging getting stuck on this axiom, because
propositional equality proofs get erased, as we described in Section
\ref{sec:semantics-of-staging}.
